---
title: "crime_regression"
author: "Paul Bradshaw"
date: "26/07/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using simple linear regression with crime data

We are going to import some data from the BBC Data Unit's [GitHub repo for a story on unsolved crime](https://github.com/BBC-Data-Unit/unsolved-crime):

```{r}
dataset = read.csv('https://raw.githubusercontent.com/BBC-Data-Unit/unsolved-crime/master/outcomes_by_force.csv')
```

This has a number of columns that can be used to generate a regression model for one of the other columns.

First we need to split the dataset into a training set and a test set:

```{r}
# Splitting the dataset into the Training set and Test set
# install.packages('caTools')
library(caTools)
#Create a seed so results are consistent with other people using same seed
set.seed(123)
#Split the dataset into a training set (0.75) and a test set (0.25)
split = sample.split(dataset$Investigation.complete..no.suspect.identified, SplitRatio = 0.75)
training_set = subset(dataset, split == TRUE)
test_set = subset(dataset, split == FALSE)
```

Now let's make a regressor which seeks to link 'no suspect identified' (the dependent variable) to 'action is not in public interest' (independent variable):

```{r}
# Fitting Simple Linear Regression to the Training set
# Specify dependent ~ independent variables, and data (training because we need test separate)
regressor <- lm(formula = Investigation.complete..no.suspect.identified ~ Formal.action.is.not.in.the.public.interest,
                data = training_set)
```

And see a summary:

```{r}
#Use summary() to see details about this model including the coefficient
#Note the dot in this:
#Formal.action.is.not.in.the.public.interest     16.12       8.52   1.892   0.0686 .  
#Likewise p-value: 0.06857
summary(regressor)
```
Some things to point out here: *2 observations deleted due to missingness* - this will be where there are `NA` values. That's quite useful - we should have made sure not to incude NA values otherwise.

Also note: 

`Formal.action.is.not.in.the.public.interest     16.12       8.52   1.892   0.0686 .`

The `.` at the end corresponds to a significance of 0.05 in the legend underneath that line.

Now let's show that line from the training set:

```{r}
# Predicting the Test set results
y_pred = predict(regressor, newdata = test_set)

# Visualising the Training set results
library(ggplot2)
ggplot() +
  geom_point(aes(x = training_set$Formal.action.is.not.in.the.public.interest, y = training_set$Investigation.complete..no.suspect.identified),
             colour = 'red') +
  geom_line(aes(x = training_set$Formal.action.is.not.in.the.public.interest, y = predict(regressor, newdata = training_set)),
            colour = 'blue') +
  ggtitle('Salary vs Experience (Training set)') +
  xlab('Years of experience') +
  ylab('Salary')
```

And see how well that line fits the test set, which wasn't included in the training calculation:

```{r}

# Visualising the Test set results
library(ggplot2)
ggplot() +
  geom_point(aes(x = test_set$Formal.action.is.not.in.the.public.interest, y = test_set$Investigation.complete..no.suspect.identified),
             colour = 'red') +
  geom_line(aes(x = training_set$Formal.action.is.not.in.the.public.interest, y = predict(regressor, newdata = training_set)),
            colour = 'blue') +
  ggtitle('Salary vs Experience (Test set)') +
  xlab('Years of experience') +
  ylab('Salary')
```

There's not a lot of data here, however, so we can try this on a larger dataset of school finances - see the relevant Rmd file in this folder for more.