---
title: "schools_regression"
author: "Paul Bradshaw"
date: "26/07/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using simple linear regression with school finance data

We are going to import some data on [academy finances](https://www.gov.uk/government/statistics/income-and-expenditure-in-academies-in-england-2014-to-2015). This has been placed in the same folder as this code, then one of the sheets copied out, then saved as csv and renamed to make it easier:

We're going to use the `tidyverse` package and `read_csv` from that to pick up any potential problems in the data formatting. Weirdly, it pulls out some headings (like "PFI Charges") and repeats them on a first row under the headings.

```{r}
library(tidyverse)
rawdatasats = read_csv('rawdatasats.csv')
```

We might want to remove that:

```{r}
#The -1 indicates every row apart from the first one
rawdatasats <- rawdatasats[-1,]
head(rawdatasats)
```


The data has a number of columns that can be used to generate a regression model for one of the other columns.

Let's rename a couple to make it easier:

```{r}
colnames(rawdatasats)[36]
colnames(rawdatasats)[36] <- "supplystaffcost"
colnames(rawdatasats)[36]
colnames(rawdatasats)[55]
colnames(rawdatasats)[55] <- "stafftraining"
colnames(rawdatasats)[55]
```

And let's check that. Now, we realise it's a character column, so we can use `parse_integer` to convert that - and then multiply it by 1000 because we know that this column refers to *thousands* of pounds.

```{r}
#summarise the staff cost column
summary(rawdatasats$supplystaffcost)
#parse it into a new vector as integers
supplystaffcost.num <- parse_integer(rawdatasats$supplystaffcost)
#summarise again, while multiplying by 1,000
summary(supplystaffcost.num*1000)
#reassign to the original column
rawdatasats$supplystaffcost <- supplystaffcost.num*1000
summary(rawdatasats$supplystaffcost)
#Now to repeat for the other column
summary(rawdatasats$stafftraining)
#These are numbers so we can leave as is.
```


First we need to split the dataset into a training set and a test set:

```{r}
# Splitting the dataset into the Training set and Test set
# install.packages('caTools')
library(caTools)
#Create a seed so results are consistent with other people using same seed
set.seed(123)
#Split the dataset into a training set (0.8) and a test set (0.2)
split = sample.split(rawdatasats$supplystaffcost, SplitRatio = 0.75)
training_set = subset(rawdatasats, split == TRUE)
test_set = subset(rawdatasats, split == FALSE)
```

Now let's make a regressor which seeks to link 'supplystaffcost' (the dependent variable) to 'stafftraining' (independent variable):

```{r}
# Fitting Simple Linear Regression to the Training set
# Specify dependent ~ independent variables, and data (training because we need test separate)
regressor <- lm(formula = supplystaffcost ~ stafftraining,
                data = training_set)
```

And see a summary:

```{r}
#Use summary() to see details about this model including the coefficient
#Note the *** in this:
#stafftraining   1572.7      155.3   10.13   <2e-16 ***
#Likewise the low p-value: < 2.2e-16
summary(regressor)
```
This is quite a strong relationship - which we should have expected. Remember that we have a bunch of NA values, however, which we should remove.

Now let's show that line from the training set:

```{r}
# Predicting the Test set results
y_pred = predict(regressor, newdata = test_set)

# Visualising the Training set results
library(ggplot2)
ggplot() +
  geom_point(aes(x = training_set$stafftraining, y = training_set$supplystaffcost),
             colour = 'red') +
  geom_line(aes(x = training_set$stafftraining, y = predict(regressor, newdata = training_set)),
            colour = 'blue') +
  ggtitle('Supply staff costs vs training costs (Training set)') +
  xlab('Training') +
  ylab('Supply costs')
```

And see how well that line fits the test set, which wasn't included in the training calculation:

```{r}

# Visualising the Test set results
library(ggplot2)
ggplot() +
  geom_point(aes(x = test_set$stafftraining, y = test_set$supplystaffcost),
             colour = 'red') +
  geom_line(aes(x = training_set$stafftraining, y = predict(regressor, newdata = training_set)),
            colour = 'blue') +
  ggtitle('Supply staff costs vs training costs (Test set)') +
  xlab('Training') +
  ylab('Supply costs')
```

There's a lot of clustering here so it's hard to see.