---
title: "R Notebook"
output: html_notebook
---

# Teacher vacancies: mapping hex tiles in R

This notebook takes [a tutorial by Rob Grant on mapping migration](http://rforjournalists.com/2017/08/09/internal-migration-part-ii-homing-in-on-individual-authorities/), and applies it to some teacher vacancy data from a BBC story.

## Compile your dataset

Let's grab the data

```{r}
#Using readr as it's supposed to be better
install.packages("readr")
library(readr)
teachervacs <- read_csv("https://raw.githubusercontent.com/BBC-Data-Unit/teacher-vacancies-deprivation/master/teacher_vacancies_by_LA_primaryLACODE.csv")
summary(teachervacs)
```

Now create a new column with the hexadecimal codes

```{r}
teachervacs$hexcode <- ifelse(teachervacs$`Average of Full-Time Posts Vacant (%)`<0.8,"#00ff00","#ff0000")
```

## Grab the shapefile for the hexagonal tiles

Any map involving shapes requires a 'shapefile': this is a set of data describing the coordinates of different shapes (countries, regions, administrative areas, etc.). In the case of hex tiles, those can be found in the [hexagonal map of local authorities](https://www.arcgis.com/home/item.html?id=593037bc399e460bb7c6c631ceff67b4) that Rob links to.

Download that zip file, unzip it, and then move all the files into the same location as this file. Now you need to get them into R - and for that we'll need some packages.


## Import mapping libraries

Here are the 3 libraries:

```{r}
install.packages("rgdal")
library(rgdal)
#sp is a dependency of rgdal so doesn't need installing
library(sp)
install.packages("rgeos")
library(rgeos)
```

To get that shape file into R we use the `readOGR` function, which is [from the `rgdal` library](https://www.rdocumentation.org/packages/rgdal/versions/1.2-8/topics/readOGR).

```{r}
#Use the name of the files without the file extension
hex <- readOGR(".","GB_Hex_Cartogram_LAs")
```


## Merge the shape file and the data frame

Now the two files are merged with `merge`. In the shape file data the authority is in a column called `LAD12CD` - we can replace `InLA` with the name of whatever our authority code column is.

```{r}
#duplicateGeoms = TRUE is used because there is an error thrown otherwise
vacmap <- merge(hex, teachervacs, by.x = "LAD12CD", by.y = "new_la_code", duplicateGeoms = TRUE)
```

## Plot the resulting merged file

Now we have a merged file (variable) we can plot that. The first argument is the name of that variable. The colour is specified with the argument `col =` followed by the colour column. Title and caption can be added with `main =` and `sub =`. The `lwd` parameter specifies the thickness of the borders.

```{r}
#the variable is called n and the column is m$color
plot(vacmap, col = vacmap$hexcode, lwd = 0.5, main = "This is the map heading", sub = "Red = some random colour\nGreen = another random colour")
```

## Export as image

If you want to use that image, you need to create a PNG 'device' using `png()`, then generate the `plot()`, and then turn it off with `dev.off()`.

```{r}
png(filename="map.png")
plot(n, col = n$color, lwd = 0.5, main = "This is the map heading", sub = "Red = some random colour\nGreen = another random colour")
dev.off()
```


## More on hex maps

RPubs maintains a [page on tilegrams in R](http://rpubs.com/bhaskarvk/tilegramsR)

ODILeeds has a [tool for creating HexJSON](https://odileeds.org/projects/hexmaps/constituencies.html)

Tony Hirst has written about creating a [HexJSON HTMLWidget for R, starting here](https://blog.ouseful.info/2017/06/28/hexjson-htmlwidget-for-r-part-1/), with two subsequent parts.
