---
title: "R Notebook"
output: html_notebook
---

# Mapping hex tiles in R 

This notebook takes [a tutorial by Rob Grant on mapping migration](http://rforjournalists.com/2017/08/09/internal-migration-part-ii-homing-in-on-individual-authorities/), and boils it down to just the mapping part. You will need:

1. Data containing local authority codes and hexadecimal colours
2. Shapefile data for local authorities and hexagonal tiles
3. Libraries for dealing with shapefiles
4. Merge the dataset and the shapefile data
5. Plot the resulting merged file

## Compile your dataset

Creating a hex tile map means we need our dataset (the data that's going to be mapped) to match the names of the hex tiles. Those are local authority codes like E06000009. 

If your dataset doesn't use these, you'll need to find one that does, and match it with yours (assuming they share another aspect, like authority name, or a different type of code)

You will also need a column full of hexadecimal codes, which will be used to colour each tile in the final map.

Again, chances are you'll need to generate those somehow. For example, by choosing 3-5 colours to represent different 'buckets' of values (lowest quartile, highest quartile, and so on) and assigning them using a formula like `IF`. Here's an example formula which uses two `IF` functions (one nested in another) to test if a value is below 33, above 66, or neither (i.e. between 33 and 66):

`=IF(C5<33,"#FF0000",IF(C5>66,"#00FF00","#0000FF"))`

For the sake of convenience I've created a dataset that you can map straight away. It's called 'tomap.csv' and it's in the same folder as this tutorial.


```{r}
#Using readr as it's supposed to be better
install.packages("readr")
testmapdata <- read.csv("tomap.csv")
```

## Find a shapefile for the hexagonal tiles

Any map involving shapes requires a 'shapefile': this is a set of data describing the coordinates of different shapes (countries, regions, administrative areas, etc.). In the case of hex tiles, those can be found in the [hexagonal map of local authorities](https://www.arcgis.com/home/item.html?id=593037bc399e460bb7c6c631ceff67b4) that Rob links to.

Download that zip file, unzip it, and then move all the files into the same location as this file. Now you need to get them into R - and for that we'll need some packages.


## Import mapping libraries

Here are the 3 libraries:

```{r}
install.packages("rgdal")
library(rgdal)
#sp is a dependency of rgdal so doesn't need installing
library(sp)
install.packages("rgeos")
library(rgeos)
```

To get that shape file into R we use the `readOGR` function, which is [from the `rgdal` library](https://www.rdocumentation.org/packages/rgdal/versions/1.2-8/topics/readOGR).

```{r}
#Use the name of the files without the file extension
hex <- readOGR(".","GB_Hex_Cartogram_LAs")
```


## Merge the shape file and the data frame

Now the two files are merged with `merge`. In the shape file data the authority is in a column called `LAD12CD` - we can replace `InLA` with the name of whatever our authority code column is.

```{r}
#variable name changed to n, merging 'testmapdata' and using new 'lacode' heading
n <- merge(hex, testmapdata, by.x = "LAD12CD", by.y = "lacode")
```

## Plot the resulting merged file

Now we have a merged file (variable) we can plot that. The first argument is the name of that variable. The colour is specified with the argument `col =` followed by the colour column. Title and caption can be added with `main =` and `sub =`. The `lwd` parameter specifies the thickness of the borders.

```{r}
#the variable is called n and the column is m$color
plot(n, col = n$color, lwd = 0.5, main = "This is the map heading", sub = "Red = some random colour\nGreen = another random colour")
```

## Export as image

If you want to use that image, you need to create a PNG 'device' using `png()`, then generate the `plot()`, and then turn it off with `dev.off()`. 

```{r}
png(filename="map.png")
plot(n, col = n$color, lwd = 0.5, main = "This is the map heading", sub = "Red = some random colour\nGreen = another random colour")
dev.off()
```


## More on hex maps

RPubs maintains a [page on tilegrams in R](http://rpubs.com/bhaskarvk/tilegramsR)

ODILeeds has a [tool for creating HexJSON](https://odileeds.org/projects/hexmaps/constituencies.html)

