---
title: "Finding destinations"
author: "Dan Wainwright and Paul Bradshaw"
date: "8 October 2018"
output: html_notebook
---

For this experiment we need data from the ons. download the zipped folders for part one and two and 
Then copy the csvs into a folder on your computer.

https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/migrationwithintheuk/datasets/internalmigrationbyoriginanddestinationlocalauthoritiessexandsingleyearofagedetailedestimatesdataset

If you don't already have them then uncomment the following two lines ("uncomment" is the term for removing the hashtag, 
anything "commented out" does not run).


```{r}
#install.packages("tidyverse")
#install.packages("dplyr")
```

#Load your libraries

```{r}
library(tidyverse)
library(dplyr)
```

#Import your two datasets



```{r}
data1 <- read_csv("Detailed_Estimates_2017_Dataset_1.csv")
data2 <- read_csv("Detailed_Estimates_2017_Dataset_2.csv")
```

#Merge your two datasets

```{r}
fulldata <- rbind(data1, data2)
#Remove the two parts
rm(data1, data2)
```


#How to filter

We need to write = twice


```{r}
youngest <- fulldata %>%
  filter(Age==18)
head(youngest)
```

For our other age groups, we need to specify each age we want

```{r}
youth <-  filter(fulldata, Age == 18 | Age == 19 | Age == 20 | Age == 21 | Age == 22 | Age == 23 | Age == 24)
older <-  filter(fulldata, Age == 25 | Age == 26 | Age == 27 | Age == 28 | Age == 29 | Age == 30)
head(youth)
head(older)
```

Now we can export our data as three csvs.

```{r}
write_csv(youngest, "youngest.csv")
write_csv(youth, "youth.csv")
write_csv(older, "older.csv")
```
#Pivot tables

We also want a unique list of LA codes for lookups in Excel, so we use R's version of a pivot table, group_by

```{r}
pivot <- fulldata %>%
  group_by(OutLA) %>%
  summarise(count=n()) %>% 
  arrange(desc(count))

write_csv(pivot, "LAs.csv")
```

Now let's narrow that further:

```{r}
top10out <- head(pivot, 10)
top10out
```

We can use this to generate a list to filter the original data against.

```{r}
#Note the double [[]] as an index - this forces the results to be a list
for (i in top10out[[1]]){print(i)}
```

Then generate the subset using `%in%` to specify we want rows where the value is in our list:

```{r}
head(fulldata)
#Note the double [[]] as an index - this forces the results to be a list
just10out <- subset(fulldata, fulldata$OutLA %in% top10out[[1]])
#Check the values
table(just10out$OutName)
```

Let's try to loop through each of these codes and find the top destination for each:

```{r}
emptyframe <- data.frame(x = character, y = numeric, z = character)
#Note the double [[]] as an index - this forces the results to be a list
for (i in top10out[[1]]){
  print(i)
  just1area <- subset(just10out, just10out$OutLA %in% i)
  #Pivot by target area by total moves 
  just1pivot <- just1area %>%
    group_by(InName) %>%
    summarise(count=n()) %>% 
    arrange(desc(count))
  just1pivot$OutCode <- i
  print(head(just1pivot))
  #Rename the columns to match what we've just grabbed
  colnames(emptyframe) <- colnames(just1pivot)
  #Add the new rows to that empty frame - we only grab the top 11 (top will be NA so it's the top 10 plus the NA)
  emptyframe <- rbind(emptyframe,head(just1pivot,11))
}

```
