---
title: "internalmigration"
author: "Paul Bradshaw"
date: "03/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Break down the problem

First:

* Import the 2 data files
* Join them into one

We want to get the picture of where young people move from and to, and where they stay put. We need to define 'young': there are 3 categories:

* 18
* 18-24
* 23-30

So we need to:

* Add a T/F column for each of those, to help with subsetting.

We also need to:

* Convert the ONS codes to actual names. Or do we?

## Install the packages

```{r}
#This first line is commented out because the package is already installed on my version of R
#install.packages("tidyverse")
library(tidyverse)
```

## Upload the data

```{r}
#read.csv is slower than read_csv for large files
pt1 <- read.csv("Detailed_Estimates_2017_Dataset_1.csv")
pt2 <- read.csv("Detailed_Estimates_2017_Dataset_2.csv")
pt_1 <- read_csv("Detailed_Estimates_2017_Dataset_1.csv")
pt_2 <- read_csv("Detailed_Estimates_2017_Dataset_2.csv")
```

We've used both `read.csv` and `read_csv` so we can compare the results:

```{r}
head(pt2)
head(pt_2)
summary(pt2)
summary(pt_2)
```

Note that one dataset (the one using `read.csv`) uses factors, while another (using `read_csv`) uses characters.

Let's remove some:

```{r}
#rm(pt_2,pt_2)
rm(pt1,pt2)
```



## Combine the data

```{r}
#fulldata <- bind_rows(pt1,pt2)
fulldata <- bind_rows(pt_1,pt_2)
```

## Summarise

```{r}
summary(fulldata)
```

## Lookup authority names

We've got a file in this folder with the LA codes and corresponding authority names - it's an .xls file so we need `read_excel`:

```{r}
library(readxl)
lalookupfull <- read_excel("lalookup.xls", skip=14)
#Now to simplify
lalookup <- lalookupfull[c(1,3)]
```

...But actually that data isn't clean - there are 4 rows that have 2 codes in them, so they won't work in a lookup. Instead we need to clean that (in this case we've cleaned it in Excel) and import the clean version:

```{r}
lalookupfull <- read_csv("lalookupcleaned.csv")
#Now to simplify
lalookup <- lalookupfull[c(1,4)]
```


Now to combine it:

```{r}
#We could instead specify what to join on
fulldata <- left_join(fulldata,lalookup, by = c("OutLA" = "Code"))
colnames(fulldata)[6] <- "OutName"
colnames(fulldata)
fulldata <- left_join(fulldata,lalookup, by = c("InLA" = "Code"))
colnames(fulldata)
colnames(fulldata)[7] <- "InName"
```



## Add category data

We can create 3 new columns in our data table by generating a vector of TRUE/FALSE answers to particular questions, and assigning that vector to a new column:

```{r}
fulldata$is18 <- fulldata$Age == 18
fulldata$is18to24 <- fulldata$Age > 17 & fulldata$Age < 25
fulldata$is23to30 <- fulldata$Age > 22 & fulldata$Age < 31
```

## Create subsets

Now let's create some subsets:

```{r}
migration18 <- subset(fulldata, fulldata$is18 == TRUE)
migration18to24 <- subset(fulldata, fulldata$is18to24 == TRUE)
migration23to30 <- subset(fulldata, fulldata$is23to30 == TRUE)
```

## Create aggregate figures for each group

And generate a list of total moves from particular authorities:

```{r}
head(table(migration18$OutName))
#Pivot by OutLA, sum of moves
outmoves18 <- tapply(migration18$Moves, migration18$OutName, sum)
outmoves18to24 <- tapply(migration18to24$Moves, migration18to24$OutName, sum)
outmoves23to30 <- tapply(migration23to30$Moves, migration23to30$OutName, sum)
#Order the data in descending numerical order
outmoves18 <- outmoves18[order(-outmoves18)]
outmoves18to24 <- outmoves18to24[order(-outmoves18to24)]
outmoves23to30 <- outmoves23to30[order(-outmoves23to30)]
write.csv(outmoves18, "outmoves18.csv")
write.csv(outmoves18to24, "outmoves18to24.csv")
write.csv(outmoves23to30, "outmoves23to30.csv")
```

The same for inbound authorities:

```{r}
#Pivot by InLA, sum of moves
inmoves18 <- tapply(migration18$Moves, migration18$InName, sum)
inmoves18to24 <- tapply(migration18to24$Moves, migration18to24$InName, sum)
inmoves23to30 <- tapply(migration23to30$Moves, migration23to30$InName, sum)
#Order the data in descending numerical order
inmoves18 <- inmoves18[order(-inmoves18)]
inmoves18to24 <- inmoves18to24[order(-inmoves18to24)]
inmoves23to30 <- inmoves23to30[order(-inmoves23to30)]
write.csv(inmoves18, "inmoves18.csv")
write.csv(inmoves18to24, "inmoves18to24.csv")
write.csv(inmoves23to30, "inmoves23to30.csv")

```


This would all be better stored in a function that we could call for different age group data tables.

## Draw a datatable

While we are at it, let's create a datatables interface in case we want to search for codes...

```{r}
DT::datatable(lalookup)
```


## Focus on the top 10 areas

We want to focus on the top few areas if we're going to draw a Sankey...

```{r}
head(outmoves18, 10)
```


