---
title: "Sankeys for internal migration"
author: "Paul Bradshaw"
date: "9 October 2018"
output: html_notebook
---

Adapted from [Creating custom Sankey diagrams using R](https://www.r-bloggers.com/creating-custom-sankey-diagrams-using-r/). Also worth seeing [Using networkD3 in R to create simple and clear Sankey diagrams](https://towardsdatascience.com/using-networkd3-in-r-to-create-simple-and-clear-sankey-diagrams-48f8ba8a4ace).

The series of steps:

* Create a dataset in the shape needed by Sankey (nodes, links)
* Generate the Sankey

## Creating a subset for top 10 areas for outward migration

In the 'findingdestingations.Rmd' notebook we generated a data frame for the top 10 authorities when it comes to people leaving:

```{r}
top10out
```

We can use this to generate a list to filter the original data against.

```{r}
#Note the double [[]] as an index - this forces the results to be a list
for (i in top10out[[1]]){print(i)}
```

Then generate the subset using `%in%` to specify we want rows where the value is in our list:

```{r}
head(fulldata)
#Note the double [[]] as an index - this forces the results to be a list
just10out <- subset(fulldata, fulldata$OutLA %in% top10out[[1]])
#Check the values
table(just10out$OutName)
```
## Getting the data in a shape for Sankey

We just need three of the columns - but there are 10:

```{r}
colnames(just10out)
```

The 3 we need are columns 5-7, but we need them in a different order: from-to-amount:

```{r}
head(just10out[c(6,7,5)])
```

Let's strip it back to those:

```{r}
just10out <- just10out[c(6,7,5)]
```


## Generate the Sankey

```{r}
# format in prep for sankey diagram
results <- tidyr::gather(results, result, vote, -Region)
# create nodes dataframe
regions <- unique(as.character(results$Region))
nodes <- data.frame(node = c(0:13), 
                    name = c(regions, "Leave", "Remain"))
#create links dataframe
results <- merge(results, nodes, by.x = "Region", by.y = "name")
results <- merge(results, nodes, by.x = "result", by.y = "name")
links <- results[ , c("node.x", "node.y", "vote")]
colnames(links) <- c("source", "target", "value")
```

