---
title: "Sankeys for internal migration"
author: "Paul Bradshaw (after Tony Hirst)"
date: "8 October 2018"
output: html_notebook
---

Most of this is adapted from [Tony Hirst's notebook](https://psychemedia.github.io/parlihacks/sankeyFlow.html):

The series of steps:

* Create a dataset in the shape needed by Sankey (from, to, label)
* Generate the Sankey

## Creating a subset for top 10 areas for outward migration

In the 'findingdestingations.Rmd' notebook we generated a data frame for the top 10 authorities when it comes to people leaving:

```{r}
top10out
```

We can use this to generate a list to filter the original data against.

```{r}
#Note the double [[]] as an index - this forces the results to be a list
for (i in top10out[[1]]){print(i)}
```

Then generate the subset using `%in%` to specify we want rows where the value is in our list:

```{r}
head(fulldata)
#Note the double [[]] as an index - this forces the results to be a list
just10out <- subset(fulldata, fulldata$OutLA %in% top10out[[1]])
#Check the values
table(just10out$OutName)
```
## Getting the data in a shape for Sankey

We just need three of the columns - but there are 10:

```{r}
colnames(just10out)
```

The 3 we need are columns 5-7, but we need them in a different order: from-to-amount:

```{r}
head(just10out[c(6,7,5)])
```

Let's strip it back to those:

```{r}
just10out <- just10out[c(6,7,5)]
```


## Generate the Sankey

```{r}
#For use in Rmd/knitr, set the block parameter: results='asis'
install.packages("googleVis")
library(googleVis)
options(gvis.plot.tag='chart')

#Generate the Sankey diagram HTML
s=gvisSankey(just10out[,c('OutName','InName','Moves')])
#And render it
plot(s)
```

