---
title: "Finding destinations"
author: "Paul Bradshaw"
date: "9 October 2018"
output: html_notebook
---

Below is a function which will return a data frame of the top 10 destination areas for each area in a list:

```{r}
getmetop10 <- function(listofcodes){
  emptyframe <- data.frame(x = character, y = numeric, z = character)
  #Note the double [[]] as an index - this forces the results to be a list
  for (i in listofcodes){
    print(i)
    just1area <- subset(just10out, just10out$OutLA %in% i)
    #Pivot by target area by total moves 
    just1pivot <- just1area %>%
      group_by(InName) %>%
      summarise(totalmoves=sum(Moves)) %>% #sum because we want to add up totals of Moves
      arrange(desc(totalmoves))
    just1pivot$OutCode <- i
    print(head(just1pivot))
    #Rename the columns to match what we've just grabbed
    colnames(emptyframe) <- colnames(just1pivot)
    #Add the new rows to that empty frame - we only grab the top 11 (top will be NA so it's the top 10 plus the NA)
    emptyframe <- rbind(emptyframe,head(just1pivot,11))
  }
  return(emptyframe)
}
```

And you run it like this:

```{r}
top10out[[1]]
nameofnewthing <- getmetop10(top10out[[1]])
```

## Function: find the top 10 in a category

Now we need to get the top 10.

```{r}
getmetop10out <- function(thedata){
  #print(head(thedata))
  pivoted <- thedata %>%
    group_by(OutName) %>%
    summarise(count=n()) %>% 
    arrange(desc(count))
  #print(head(pivoted,10))
  return(head(pivoted,10))
}
```

And run it:

```{r}
top10outfor18s <- getmetop10out(migration18)
```

## Combining the two into one function

This is what we want:

Given a subset of internal migration by a particular category, return a list of the top 11 destinations for people leaving that area, for each of the 10 areas that most people leave.

```{r}
getmetop11oftop10 <- function(thedata){
  #first pivot the data to find the top 10
  pivoted <- thedata %>%
    group_by(OutName) %>%
    summarise(count=n()) %>% 
    arrange(desc(count))
  #print(head(pivoted,10))
  top10forthis <- head(pivoted,11)
  #Now for the second part
  listofcodes <- top10forthis[[1]]
  emptyframe <- data.frame(x = character, y = numeric, z = character)
  #Note the double [[]] as an index - this forces the results to be a list
  for (i in listofcodes){
    print(i)
    #subset to just those moves involving the top 10 names
    just1area <- subset(thedata, thedata$OutName %in% i)
    #Pivot by target area by total moves 
    just1pivot <- just1area %>%
      group_by(InName) %>%
      summarise(totalmoves=sum(Moves)) %>% #sum because we want to add up totals of Moves
      arrange(desc(totalmoves))
    just1pivot$OutCode <- i
    #print(head(just1pivot))
    #Rename the columns to match what we've just grabbed
    colnames(emptyframe) <- colnames(just1pivot)
    #Add the new rows to that empty frame - we only grab the top 11 (top will be NA so it's the top 10 plus the NA)
    emptyframe <- rbind(emptyframe,head(just1pivot,11))
  }
  return(emptyframe)
}
```

Now to test:

```{r}
migration18.top110 <- getmetop11oftop10(migration18)
migration18to24.top110 <- getmetop11oftop10(migration18to24)
migration23to30.top110 <- getmetop11oftop10(migration23to30)
```

May as well export the results...

```{r}
write.csv(migration18.top110, "migration18dest.csv")
write.csv(migration18to24.top110, "migration1824dest.csv")
write.csv(migration23to30.top110, "migration2330dest.csv")
```

