---
title: "Fetching data for crimes near airports"
author: "Paul Bradshaw"
date: "24/7/2018"
output: html_document
---

# Fetching data for crimes near airports

For this project we need to break down the challenge first:

1. Get crimes for one airport and one month
2. Repeat for all months over a period (3 years, for example)
3. Collate locations of multiple airport car parks
4. Repeat for those locations

## Using the Data.police.uk API in R

The [data.police.uk API](https://data.police.uk/docs/) provide a range of data related to police forces, crimes and stop and search in the UK. Data begins in 2010 and the most recent data tends to be about 3 months old.

## Grabbing data on crimes at a location

[The documentation for the 'method' to grab data on crimes at a particular location is here](https://data.police.uk/docs/method/crime-street/).

Data is fetched by forming a URL that contains your question. For example, the example given of an API call is [https://data.police.uk/api/crimes-street/all-crime?lat=52.629729&lng=-1.131592&date=2013-01](https://data.police.uk/api/crimes-street/all-crime?lat=52.629729&lng=-1.131592&date=2013-01)

This includes:

* A type of crime: `all-crime?`
* A latitude: `lat=`
* A longitude: `lng=`
* And a month: `date=`

Note that you cannot ask for a year: the date has to be a year *and* month. If you don't specify a date, then it appears to provide the most recent month for which it holds data.

Note also that **locations are anonymised to the nearest location from a list**. In other words, the data does not mean that crimes took place at this particular location, but somewhere in this vicinity. 

### Crimes near Manchester Airport

A search for Manchester Airport on Google Maps brings up a lat-long of 53.3588026, -2.2727303. This is approximate for now - we'll look for more specific locations later but just need an address to test.

This can be added to the data.police.uk query URL for 'crimes as a particular location', along with the latest month for which data is available (May 2018) like so:

`https://data.police.uk/api/crimes-at-location?date=2018-05&lat=53.3588026&lng=-2.2727303`

```{r}
manurl <- "https://data.police.uk/api/crimes-at-location?date=2018-05&lat=53.3588026&lng=-2.2727303"
```


Now we need to fetch data from that.

## Working with the JSON

To convert JSON data into a data variable that R can work with, we use the `jsonlite` library ([documentation here](https://cran.r-project.org/web/packages/jsonlite/jsonlite.pdf)). This should already be installed in RStudio (if not, type `install.packages('jsonlite')`), so you just need to activate it.

```{r}
library('jsonlite')
```

Once added to your library, we use the `fromJSON` function to import JSON data from a URL into a new variable like so:

```{r}
mancrimes <-fromJSON(manurl)
#And summarise the category column
table(mancrimes$category)
```

It's a good idea to have the URL open in a browser at the same time so you can see the structure and work out how to access the bit you're after. You should use Chrome or Firefox with the extension [JSONView](https://chrome.google.com/webstore/detail/jsonview/chklaanhfefbnpoihckbnefhakgolnmc?hl=en) installed, as this makes it a lot easier to understand.

Because the JSON is a list of pairs, the result is created as a table (other APIs may generate different types of results)

## Generate URLs for other dates

Let's try to grab the data for the same month in previous years. The police recently decided to only provide 3 years' worth of data, so there's no results for May 2015 or earlier years (the last month available is June 2015).

```{r}
#create a sequence from 2016 to 2018 and store in variable 'years'
years <- seq(2016,2018)
years
```

Once created, we can generate URLs by looping through those numbers and adding them to a base url, using a `for` loop:

```{r}
#create an empty vector (list) to store the URLs generated
airportyearurls <- c()
#To make things easier we can move the date= parameter to the end, but leave it empty:
parturl <- "https://data.police.uk/api/crimes-at-location?lat=53.3588026&lng=-2.2727303&date="
#create a series of URLs using those numbers
for(year in years){
  #test loop is working
  #use paste to concatenate the year number with the base url, and no separator - store in 'airportyearurl' variable
  airportyearurl <- paste(parturl,year,"-05",sep="")
  #print that variable
  print(airportyearurl)
  #combine this monthurl with the list of all monthurls so far
  airportyearurls <- c(airportyearurls,airportyearurl)
}
airportyearurls
```

## Grab the data from the URLs

Now to loop through each of those URLs and grab the data...

```{r}
#We need to activate the dplyr library to use bind_rows below
library(dplyr)
#Use our test dataset to create the empty data frame (empty because we specify 0 rows)
#This is needed because we need to have the same number of columns as the data we will add next
airportdatafull <- head(mancrimes,0, row.names=NULL)
#loop through each url and store it in 'i' variable
#use [1:1] to limit it to first item
for(i in airportyearurls){
  print(i)
  airport1yr <- fromJSON(i, flatten=TRUE) #flatten = TRUE ensures no problems with nested branches
  #bind_rows is the dplyr way of combining the data, best when columns might not match
  #airportallyr.bindrows <- bind_rows(airport1yr,airportdatafull)
  #rbind is another way of doing the same, but columns must match
  airportdatafull <- rbind(airportdatafull, airport1yr)
}
#delete the temporary variable
rm(airport1yr)
#remove mancrimes which is was only 1 month and is now superceded by the new data
rm(mancrimes)
#Check all years have been grabbed
table(airportdatafull$month)
```

There is an increase here - it doesn't look big but it's still 10% since 2017 and 5% since 2016. We need more months and locations to get a better picture, but we're getting closer.

Given that we know only the last 3 years - or 36 months - of data is available, we can extend our loop to factor that in.

```{r}
#Set our start and end points
latestyr <- 2018
latestmonth <- 5
earliestyr <- latestyr-3
earliestmonth <- latestmonth+1
#create an empty vector (list) to store the URLs generated
airportmonthurls <- c()
#To make things easier we can move the date= parameter to the end, but leave it empty:
parturl <- "https://data.police.uk/api/crimes-at-location?lat=53.3588026&lng=-2.2727303&date="
#create a series of URLs using those numbers
for(year in seq(earliestyr,latestyr)){
  #test loop is working
  #use paste to concatenate the year number with the base url, and no separator 
  #store in airportyearurl variable
  for (month in months){
    if(year == earliestyr & month < earliestmonth){
      print("too early to provide data")
    }
    else if(year == latestyr & month > latestmonth){
      print("too late to provide data")
    }
    else
    {
      airportmonthurl <- paste(parturl,year,"-",month,sep="")
      #print that variable
      print(airportmonthurl)
      #combine this monthurl with the list of all monthurls so far
      airportmonthurls <- c(airportmonthurls,airportmonthurl)
    }
  }
}
```



We should at the end have 36 URLs for the latest 36 months. Note that you will need to update the `latestyear` and `latestmonth` variables if you run this again after new data has been added.

Now to grab the data from those URLs:

```{r}
#reset the dataframe so it only contains the headers
airportdatafull <- head(airportdatafull,0, row.names=NULL)
#loop through each url and store it in 'i' variable
#use [1:1] to limit it to first item
for(i in airportmonthurls){
  #print(i)
  airport1month <- fromJSON(i, flatten=TRUE) #flatten = TRUE ensures no problems with nested branches
  #bind_rows is the dplyr way of combining the data, best when columns might not match
  #airportallyr.bindrows <- bind_rows(airport1yr,airportdatafull)
  #rbind is another way of doing the same, but columns must match
  airportdatafull <- rbind(airportdatafull, airport1month)
}
#Check all years have been grabbed
table(airportdatafull$month)
```

Now let's store those results in a data frame:

```{r}
crimeslast3yrs <- data.frame(table(airportdatafull$month))
write.csv(crimeslast3yrs,"crimes3yrs.csv")
write.csv(airportdatafull,"fullcrimedata.csv")
```

We can filter it a little too:


