---
title: "Analysing stop and search data"
output: html_notebook
---

# Analysing stop and search data

First we need to plan the steps in the project. These are:

1. Grab data from the API
2. Loop through each force
3. Loop through each month
4. Combine datasets into one
5. Analysis: change over time
6. Analysis: 

## Grab data from the API

The Police API [provides data on stop and search](https://data.police.uk/docs/method/stops-force/). An example URL for data is [https://data.police.uk/api/stops-force?force=avon-and-somerset&date=2017-01](https://data.police.uk/api/stops-force?force=avon-and-somerset&date=2017-01).

We need to grab the data from that URL and convert to a data frame.

```{r}
#store test url in a variable 'testurl'
testurl <- "https://data.police.uk/api/stops-force?force=avon-and-somerset&date=2017-01"

#to deal with JSON we need the jsonlite package
library(jsonlite)

#to convert from JSON to a data frame we use the function fromJSON
#here we store the results in a variable called 'testdata'
testdata <- fromJSON(testurl, flatten=TRUE) #some parts have further branches, so flatten=TRUE ensures these are flattened to the same level - this results in 20 variables rather than 16
```

## Looping through multiple months

First we need to create a list of months.

To create a range of numbers in R you can [use the `seq` (sequence)](http://stat.ethz.ch/R-manual/R-devel/library/base/html/seq.html) function. This needs two parameters - a start and end point - and can also take a third parameter specifying interval (e.g. every third number).

```{r}
#create a sequence from 1 to 12 and store in variable 'months'
months <- seq(1,12)
```

Once created, we can generate URLs by looping through those numbers and adding them to a base url, using a `for` loop:

```{r}
#create an empty vector (list) to store the URLs generated
monthurls <- c()
#create a series of URLs using those numbers
for(month in months){
  #test loop is working
  #print(month)
  #use paste to concatenate the month number with the base url, and no separator - store in 'monthurl' variable
  monthurl <- paste("https://data.police.uk/api/stops-force?force=avon-and-somerset&date=2017-",month,sep="")
  #print that variable
  print(monthurl)
  #combine this monthurl with the list of all monthurls so far
  monthurls <- c(monthurls,monthurl)
}
monthurls
```

## Expand to include multiple years

The earliest data from the API comes from [July 2015](https://data.police.uk/api/stops-force?force=avon-and-somerset&date=2015-07) - queries before come back empty or dead.

We need to generate a list of years from 2015 to 2018, then add months to those.

```{r}
#create empty list to store year urls
yearurls <- c()
#create empty list to store full urls
fullurls <- c()
#create vector of years from 2015-2018
years <- seq(2015,2018)
for(year in years){
  #test loop is working
  print(year)
  #use paste to concatenate the month number with the base url, and no separator - store in 'monthurl' variable
  yearurl <- paste("https://data.police.uk/api/stops-force?force=avon-and-somerset&date=",year,sep="")
  #print that variable
  print(yearurl)
  for(month in months){
    fullurl <- paste(yearurl,month,sep="-")
    print(fullurl)
    fullurls <- c(fullurls,fullurl)
  }
  #combine this monthurl with the list of all monthurls so far
  yearurls <- c(yearurls,yearurl)
}
```



## Grab data from all URLs

Now to loop through each of the URLs in the list we've just made, and fetch the JSON as we did with the test url at the start.

First, we know the first 6 URLs are dead or empty, so we need to remove those from the list

```{r}
#replace fullurls with the items in fullurls from the 7th position onwards, to remove dead links for first 6 months of 2015
fullurls <- fullurls[7:length(fullurls)]
```

Now to start the loop...

```{r}
fulldata <- data.frame()
#loop through each url and store it in 'i' variable
#use [1:1] to limit it to first item
for(i in fullurls){
  print(i)
  fulldata <- fromJSON(i, flatten=TRUE)
}
```

At the moment that loop creates a data frame and then *overwrites* it each time it loops again. Instead, we want it to *add to* the data frame.

One way to add to a data frame is to use `rbind`:

```{r}
#We need to activate the dplyr library to use bind_rows below
library(dplyr)
#Use our test dataset to create the empty data frame (empty because we specify 0 rows)
#This is needed because we need to have the same number of columns as the data we will add next
fulldata <- head(testdata,0, row.names=NULL)
#use [1:1] to limit it to first item
for(i in fullurls[5:10]){
  print(i)
  #convert from JSON
  thismonth <- fromJSON(i, flatten=TRUE) #flatten = TRUE ensures no problems with nested branches
  #bind_rows is the dplyr way of combining the data, best when columns might not match
  allmonths.bindrows <- bind_rows(thismonth,fulldata)
  #rbind is another way of doing the same, but columns must match
  allmonths.rbind <- rbind(fulldata, thismonth)
}


```

At this point we have over 13000 records of stops and searches. It's worth generating a pivot [table](https://www.r-bloggers.com/r-function-of-the-day-table/):

```{r}
table(allmonths.bindrows$age_range)
sort(table(allmonths.bindrows$self_defined_ethnicity), decreasing = T)
```

This gives us an idea of the analysis that can be done - but remember we still only have one police force. Next, we need to grab all the others...

## Creating a vector (list) of police forces

Looping through all the police forces will be a similar process to looping through months and years. However, this time we can't simply generate a range of numbers - instead we need to find a definitive list of the police force names (strings) used in the police API.

As it happens, [one of the API *methods* returns just such a list](https://data.police.uk/docs/method/forces/.

```{r}
#store the URL that has force names and ids
forcesjson <- "https://data.police.uk/api/forces"
#convert to dataframe
forcesdf <- fromJSON(forcesjson)
#extract ids only - because this is one column it is now a vector (list)
forceslist <- forcesdf$id
```

### Looping through each police force and adding it to a URL

Now we have a list we can loop through it in the same way. By now we are going to have 3 loops in total, each one nested within the other:

* Loop through each force to create a partial URL. For each...
* ...loop through each year to add that to the URL too. For each year...
* ...loop through each month to add that to the URL, completing it.

Only within that last loop does the JSON get grabbed and combined as before.

First, then, here's the code to loop through each force in that list and insert it into the URL. This time we need to add the `&date=` *after* that, as well.

```{r}
#create empty list to store year urls
yearurls <- c()
#create empty list to store full urls
fullurls <- c()
#create vector of years from 2015-2018
years <- seq(2015,2018)
#loop through forces
for (force in forceslist){
  #concatenate url inserting the force name along with other elements
  forceurl <- paste("https://data.police.uk/api/stops-force?force=",force,"&date=",sep="")
  print (forceurl)
}
```

Now we can take the code from before and place that *within* our new loop:

```{r}
#create empty list to store year urls
yearurls <- c()
#create empty list to store full urls
fullurls <- c()
#create vector of years from 2015-2018
years <- seq(2015,2018)
#loop through forces
for (force in forceslist){
  #concatenate url inserting the force name along with other elements
  forceurl <- paste("https://data.police.uk/api/stops-force?force=",force,"&date=",sep="")
  print (forceurl)
  #now the code from earlier sits inside this for loop, so it runs for each force
  for(year in years){
    #test loop is working
    print(year)
    #note this now uses the new forceurl variable
    yearurl <- paste(forceurl,year,sep="")
    #print that variable
    print(yearurl)
    for(month in months){
      fullurl <- paste(yearurl,month,sep="-")
      print(fullurl)
      fullurls <- c(fullurls,fullurl)
    }
    #combine this monthurl with the list of all monthurls so far
    yearurls <- c(yearurls,yearurl)
  
  }
}
print(length(fullurls))
```


At this point we have 2112 URLs: 44 forces times 4 years (176) times 12 months.

Remember that we've re-created the problem with the dead URLs and will need to strip those out again. This time it's not as simple as removing the first 6 - we'll need to remove the first 6 in every set of 48.


## Don't forget the stops with no location data

https://data.police.uk/api/stops-no-location?force=cleveland&date=2017-01


## TBC: Adapt the previous loop for our new list of 2112 URLs

```{r}
#We need to activate the dplyr library to use bind_rows below
library(dplyr)
#Use our test dataset to create the empty data frame (empty because we specify 0 rows)
#This is needed because we need to have the same number of columns as the data we will add next
fulldata <- head(testdata,0, row.names=NULL)
#use [1:1] to limit it to first item
for(i in fullurls[5:10]){
  print(i)
  #convert from JSON
  thismonth <- fromJSON(i, flatten=TRUE) #flatten = TRUE ensures no problems with nested branches
  #bind_rows is the dplyr way of combining the data, best when columns might not match
  allmonths.bindrows <- bind_rows(thismonth,fulldata)
  #rbind is another way of doing the same, but columns must match
  allmonths.rbind <- rbind(fulldata, thismonth)
}


```
