---
title: "shiny"
author: "Paul Bradshaw"
date: "18/05/2019"
output: html_document
---

# Starting with Shiny

Shiny is an R package for creating interactive web apps. First, you need to install/activate it:

```{r}
#Shiny comes pre-installed on modern versions of RStudio, so this line is commented out
#install.packages("shiny")
#Activate the Shiny package
library("shiny")
```

There are some built-in examples that you can run with `runExample`:

```{r}
#Show what examples we can run by leaving the parentheses empty:
shiny::runExample()
```

```{r}
#Try it with the first example
shiny::runExample("01_hello")
```

These will open in RStudio's built-in browser. Note that when you stop it, you will get a message about "Listening on http://127.0.0.1:7914". If you click on the 'open in browser' button you will see it open at 127.0.0.1:7914 too.

To run any more lines you must stop the code running above.

```{r}
find.package("shiny")
```

## Create an app

The code below shows 3 lines to create a minimal app. If you: 

* Create a new R file by selecting **File > New File > R Script** 
* Paste this code into that R file
* Save into its own directory
* Give it the name `app.R`

Then when you open it (or after it has saved) there should be a button to 'Run App' indicating that it has recognised that this is an app.

```{r, eval=FALSE, include=FALSE}
#Define the UI object
ui <- shiny::fluidPage()
#Create a function called 'server' that takes two parameters it calls input and output
server <- function(input, output) {}
#use those two as parameters for the shinyApp function
shiny::shinyApp(ui = ui, server = server)
```

You can also run this file with `runApp()` like so:

```{r}
#You only need to name the folder - because there should not be more than one app.R file in any directory
shiny::runApp("Minimal_App", launch.browser = T)
```

## Adding input/output to the app

Now we expand the minimal skeleton of the Shiny app to include some input and output:

```{r, eval=FALSE, include=FALSE}
#Define the UI object
ui <- shiny::fluidPage(
  #Add a slider to our UI, specifying various parameters
  sliderInput(inputId = "bins",
              label = "Number of bins:",
              min = 1,
              max = 50,
              value = 30), #default value
  #Add an output
  plotOutput(outputId = "hist")
)
#Create a function called 'server' that takes two parameters it calls input and output
server <- function(input, output) {}
#use those two as parameters for the shinyApp function
shiny::shinyApp(ui = ui, server = server)
```

If that's saved as a separate 'app.R' file in a directory called 'histo_app' then we can run it as follows:

```{r}
#You only need to name the folder - because there should not be more than one app.R file in any directory
shiny::runApp("histo_app", launch.browser = T)
```

```{r}
goals <- read.csv("Goals-Grid view.csv")
games <- read.csv("Games-Main View.csv")
games$na <- is.na(games$Goals.for)
games <- subset(games, games$na == FALSE)
```

## Creating an interactive table

You can use different output functions - like `dataTableOutput` - to generate other types of dashboards, including tables. The code below, saved in the 'table_app' folder, creates an interactive table to display the 'goals' data frame.

```{r}
#Define the UI object
ui <- shiny::fluidPage(
  #Add a table to our UI, specifying various parameters
  h4("Goals by the Bears"), #Set a heading
  shiny::dataTableOutput("dfHead") #this naming means it can be called as output$dfHead below
)
#Create a function called 'server' that takes two parameters it calls input and output
server <- function(input, output) {
  output$dfHead <- shiny::renderDataTable(goals) #render the goals data frame
}
#use those two as parameters for the shinyApp function
shiny::shinyApp(ui = ui, server = server)
```

