---
title: "stata election survey"
author: "Paul Bradshaw"
date: "5/2/2017"
output: html_document
---

# Reading election survey data in Stata or SPSS format

The [British Election Study publishes its data in Stata and SPSS formats](http://www.britishelectionstudy.com/data-object/post-election-wave-6-of-the-2014-2017-british-election-study-internet-panel/). 

[This overview](http://www.statmethods.net/input/importingdata.html) provides guidance on importing a number of types of data file. We'll start with Stata

```{r}
#Install the foreign library, which allows us to use the read.dta function
library(foreign)
#We can try to read it directly from the URL
mydata <- read.dta('http://www.britishelectionstudy.com/custom/uploads/2015/07/BES2015_W6_v1.2.dta')
```

Unfortunately this generates an error: `not a Stata version 5-12 .dta file`. A bit of googling around suggests this might be because the `foreign` package was [frozen at Stata 12](https://stackoverflow.com/questions/23899009/read-stata-13-file-in-r)

Instead, then, we try a different package, `readstata13`, designed to address this problem:

```{r}
install.packages("readstata13")
library(readstata13)
electionsurvey15 <- read.dta13('BES2015_W6_v1.2.dta')
```

Now to write that as a new CSV file:

```{r}
write.csv(electionsurvey15, 'BES2015_W6_v1.2.csv')
```

## Doing the same with the SPSS file

For the SPSS challenge I'm using the steps [outlined in this guide](http://www.milanor.net/blog/how-to-open-an-spss-file-into-r/). I've downloaded the file first, and placed it in the same directory as this R project:

```{r}
spssdataset = read.spss("BES2015_W6_v1.2.sav", to.data.frame=TRUE)
```

Once again you can then write that object as a CSV like so:

```{r}
write.csv(spssdataset, 'BES2015_W6_v1.2.csv')
```


## Using Haven to do both

[The Stack Overflow thread on this problem](https://stackoverflow.com/questions/23899009/read-stata-13-file-in-r/23899204) notes that "there's a new package called Haven, by Hadley Wickham, which can load Stata 13 dta files (as well as SAS and SPSS files)". However, the code outlined returns an error `"there is no package called 'haven'"`

Instead, searching for the haven package takes you to [this GitHub](https://github.com/tidyverse/haven) where you'll find new instructions for installing and using Haven, which I've adapted below:

```{r}
# The easiest way to get haven is to install the whole tidyverse:
install.packages("tidyverse")
# Alternatively, install just haven:
install.packages("haven")
library(haven)

# SPSS
bes2015spss <- read_sav("BES2015_W6_v1.2.sav")
#You can also read from a URL
bes2015spss <- read_sav("http://www.britishelectionstudy.com/custom/uploads/2015/07/BES2015_W6_v1.2.sav")
# Stata
bes2015stata <- read_dta("BES2015_W6_v1.2.dta")

#Just grabbing some cols
votes15 <- bes2015spss[c(5,8,13,15,25)]
#Find which columns refer to non registration
colstokeep <- grep("NotRegistered",colnames(bes2015spss))
```



