---
title: "dropout rates"
author: "Paul Bradshaw"
date: "8 March 2017"
output: html_document
---

## How to do it

First grab the Excel file and create a subset of the columns we want.

```{r}
install.packages('readxl')
library(readxl)
#Grab the file locally
mynewdata <- read_excel('UKPIs_2014-15_Non-continuation.xlsx', sheet=1, skip=3)
#The data is in columns 40 and 41, but we also want the name and codes in the first 3 cols, and the number of students
colswewant <- c(1,2,3,4,40,41,42,43)
#Use that to create a subset of just those columns
subset <- mynewdata[colswewant]
# Grab the notes too, which are in rows 178 onwards in the new dataset (remember we chopped out the first few rows)
notes <- c(178:188)
subset[notes,1]
```

We can grab it from a URL:

```{r}
#we need the httr library to do this
library("httr", lib.loc="/Library/Frameworks/R.framework/Versions/3.1/Resources/library")
url <- 'https://www.hesa.ac.uk/files/UKPIs_2014-15_Non-continuation.xlsx'
GET(url, write_disk("nonconts.xlsx", overwrite=TRUE))
mynewdata <- read_excel('nonconts.xlsx', sheet=1, skip=3)
#The data is in columns 40 and 41, but we also want the name and codes in the first 3 cols, and the number of students
colswewant <- c(1,2,3,4,40,41,42,43)
#Use that to create a subset of just those columns
subset <- mynewdata[colswewant]
# Grab the notes too, which are in rows 178 onwards in the new dataset (remember we chopped out the first few rows)
notes <- c(178:188)
subset[notes,1]
```

Let's give the last 2 columns easier names to use:

```{r}
colnames(subset)[5] <- "percnolongerinHE"
colnames(subset)[6] <- "benchmark"
```

Create a new column to calculate difference from benchmark:

```{r}
subset$difference <- as.numeric(subset$benchmark) - as.numeric(subset$percnolongerinHE)
#And then sort by that new column
sorted <- subset[order(subset$difference),]
```

Now to create a file with the results

```{r}
write.csv(sorted,"dropouts1415.csv")
```

# Grabbing the new year of data

```{r}
url <- '[PASTE THE XLSX URL HERE]'
GET(url, write_disk("16data.xlsx", overwrite=TRUE))
mynewdata <- read_excel('16data.xlsx', sheet=1, skip=3)
subset16 <- mynewdata[colswewant]
# Grab the notes too, which are in rows 178 onwards in the new dataset (remember we chopped out the first few rows)
notes16 <- c(178:188)
subset16[notes,1]
```

# Merging datasets

Once we've created a new dataset called `subset16` we can merge them like so:

```{r}
twosets <- merge(subset,subset16, by="UKPRN", suffixes="15")
#Add a new column showing the difference between each year's data (the difference between the differences!)
twosets$change <- twosets$difference15 - twosets$difference
twosetssorted <- twosets[order(subset$change),]
write.csv(twosetssorted,'15v16results.csv')
```

Grab the full data 

```{r}
url <- 'https://github.com/paulbradshaw/Rintro/blob/master/calculations/dropouts.xlsx?raw=true'
GET(url, write_disk("allyears.xlsx", overwrite=TRUE))
allyears <- read_excel('allyears.xlsx', sheet=1)
```

Grab the new data and add it to that, add calculations for biggest rises and falls, sort, then export

```{r}
# Fetch the data from the new URL
url <- '[PASTE THE XLSX URL HERE]'
#Create a new file in the same folder
GET(url, write_disk("16data.xlsx", overwrite=TRUE))
#Read from that new file into R
mynewdata <- read_excel('16data.xlsx', sheet=1, skip=3)
#Create a subset just containing the key columns
subset16 <- mynewdata[colswewant]
# Grab the notes too, which are in rows 178 onwards in the new dataset (remember we chopped out the first few rows)
notes16 <- c(178:188)
subset16[notes,1]
#put the perc no longer in HE into our all years dataset

```

