---
title: "Datatables"
author: "Paul Bradshaw"
date: "03/10/2018"
output: html_document
---

# Using R to create interactive datatables

This is a very short notebook that explains how to use the `DT` package to create an interactive table.

First, we need to import some data. In the same folder as this notebook is a CSV file with some data we can use:

```{r}
#Import data from the CSV file into a new data frame object called gpdata
gpdata <- read.csv("gpdata.csv")
```

There's a lot here, so let's see the first few lines:

```{r}
#Show the first few lines of gpdata
head(gpdata)
```

And get an overview of the column names:

```{r}
#Show the column names of gpdata
colnames(gpdata)
```

And let's reduce it to just a few columns

```{r}
#Show the column names of gpdata - but just those for columns 8 to 16
colnames(gpdata)[c(8:16)]
```

Then create a separate dataset with just those columns

```{r}
#Create a new data frame called gpsubset - by extracting columns 8 to 16 of gpdata
gpsubset <- gpdata[c(8:16)]
```


## Draw a datatable

Now let's create a datatables interface to look at that. To do that we use the `DT` package:

```{r}
#This line is commented out because DT is already installed
#install.packages("DT")
#This activates DT in the library, although again it isn't really needed
library(DT)
#Now use DT to generate a datatable for the data frame gpsubset
DT::datatable(gpsubset)
```

This will preview in the notebook itself, but you can also generate a HTML version of this notebook by clicking the **Knit** button above the notebook, and selecting **Knit to HTML**.
