---
title: "using_police_api"
author: "Paul Bradshaw"
date: "4/12/2017"
output: html_document
---

# Using the Data.police.uk API in R

The [data.police.uk API](https://data.police.uk/docs/) provide a range of data related to police forces, crimes and stop and search in the UK. Data begins in 2010 and the most recent data tends to be about 3 months old.

## Grabbing data on crimes at a location

[The documentation for the 'method' to grab data on crimes at a particular location is here](https://data.police.uk/docs/method/crime-street/).

Data is fetched by forming a URL that contains your question. For example, the example given of an API call is [https://data.police.uk/api/crimes-street/all-crime?lat=52.629729&lng=-1.131592&date=2013-01](https://data.police.uk/api/crimes-street/all-crime?lat=52.629729&lng=-1.131592&date=2013-01)

This includes:

* A type of crime: `all-crime?`
* A latitude: `lat=`
* A longitude: `lng=`
* And a month: `date=`

Note that you cannot ask for a year: the date has to be a year *and* month. If you don't specify a date, then it appears to provide the most recent month for which it holds data.

### Crimes within an area

You can also look for crimes within a polygon area. The example give is: 

`https://data.police.uk/api/crimes-street/all-crime?poly=52.268,0.543:52.794,0.238:52.130,0.478&date=2013-01`

In this case there are three lat-long pairs that come after `poly=`, with each separated by a colon:

* `52.268,0.543`
* `:`
* `52.794,0.238`
* `:`
* `52.130,0.478`

So this draws a polygon between 3 points (a triangle) and provides data for crimes within that triangle.

## Finding the lat-longs for a neighbourhood

You can get the lat-longs for a neighbourhood boundary by [using the method outlined here](https://data.police.uk/docs/method/neighbourhood-boundary/). An example response would be: `https://data.police.uk/api/leicestershire/NC04/boundary`

To get a list of neighbourhoods use a URL like: `https://data.police.uk/api/leicestershire/neighbourhoods`

And a list of forces at: `https://data.police.uk/api/forces`

## Working with the JSON

Now we need to do something with those URLs to extract the JSON. Some context first...

To convert JSON data into a data variable that R can work with, we use the `jsonlite` library ([documentation here](https://cran.r-project.org/web/packages/jsonlite/jsonlite.pdf)). This should already be installed in RStudio (if not, type `install.packages('jsonlite')`), so you just need to activate it.

```{r}
library('jsonlite')
```

Once added to your library, we use the `fromJSON` function to import JSON data from a URL into a new variable like so:

```{r}
policeforces <-fromJSON("https://data.police.uk/api/forces")
```

It's a good idea to have the URL open in a browser at the same time so you can see the structure and work out how to access the bit you're after. You should use Chrome or Firefox with the extension [JSONView](https://chrome.google.com/webstore/detail/jsonview/chklaanhfefbnpoihckbnefhakgolnmc?hl=en) installed, as this makes it a lot easier to understand.

Because the JSON is a list of pairs, the result is created as a table (other APIs may generate different types of results)

The same approach can be used to grab all the neighbourhoods within a particular force:

```{r}
gmneighbourhoods <- fromJSON("https://data.police.uk/api/greater-manchester/neighbourhoods")
```

To get all neighbourhoods for all forces, we'd need to:

1. Create a `for` loop which went through each force
2. Form a neighbourhoods URL with the force id in the right place
3. Store the results in a table (appending it to the results so far each time)

We would also need to grab all the lat-long pairs to describe each neighbourhood like so:

```{r}
latlongsac29 <- fromJSON("https://data.police.uk/api/greater-manchester/AC29/boundary")
```

Now to generate a URL containing all the lat-longs. Here's the example again:

* `https://data.police.uk/api/crimes-street/all-crime?poly=`
* `52.268,0.543:52.794,0.238:52.130,0.478`
* `&date=2013-01`

To generate the URL for one area we need to extract all the lat-long pairs and put them into one string, with colons between each one. The code below tests this process on one neighbourhood - later we'll need to store this as a function so it can be run multiple times on different areas.

* the `seq` function generates a range between two numbers - in this case `1` and the number of entries in the 'latitude' field in `latlongsac29`
* the `paste` function is used in two different ways: the first time it is used to combine 3 items to make one string (using `sep=""` to specify no space between each item)
* the second time it is used to combine all items in a vector to make one string, using `collapse=":"` to insert a colon between each item. We could have added a colon in the earlier `paste` command, as a string, but that would have left us with one at the end.

```{r}
#create an empty vector to hold the lat/longs
polygonlatlongs <- c()
#loop through our imported JSON - as many times as there are entries
for (i in seq(1,length(latlongsac29$latitude))) {
  #print(i)
  #put the lat-long pair together, with a comma between
  pair <- paste(latlongsac29$latitude[i],",",latlongsac29$longitude[i], sep="")
  #print(pair)
  #add the pair to the list
  polygonlatlongs <- c(polygonlatlongs,pair)
  #print(polygonlatlongs)
}
#Now the loop has finished, take all the pairs from the list and convert into a single string with a colon between each item
polygonforurl <- paste(polygonlatlongs,collapse=":")
#insert that string into the URL
polygoninurl <- paste("https://data.police.uk/api/crimes-street/all-crime?poly=",polygonforurl,"&date=2016-01",sep="")
print(polygoninurl)
```

That URL works, so we'll store this as a function, changing two elements to variables (`latlongs` and `date` need to be passed at the start) so they can be changed:

```{r}
returnpolygonurl <- function(latlongs,crimetype,date){
  #latlongs should be table containing latitude and longitude, grabbed from JSON
  #crimetype should be from list at https://data.police.uk/api/crime-categories
  #date must be in following format: "2016-01", or you can use "" to get latest month
  #create an empty vector to hold the lat/longs
  polygonlatlongs <- c()
  #loop through our imported JSON - as many times as there are entries
  for (i in seq(1,length(latlongs$latitude))) {
    #put the lat-long pair together, with a comma between
    pair <- paste(latlongs$latitude[i],",",latlongs$longitude[i], sep="")
    #add the pair to the list
    polygonlatlongs <- c(polygonlatlongs,pair)
  }
  #Now the loop has finished, take all the pairs from the list and convert into a single string with a colon between each item
  polygonforurl <- paste(polygonlatlongs,collapse=":")
  #insert that string into the URL
  polygoninurl <- paste("https://data.police.uk/api/crimes-street/",crimetype,"?poly=",polygonforurl,"&date=",date,sep="")
  print(polygoninurl)
  return(polygoninurl)
}
```

Now test it:

```{r}
#store the resulting URL in a new variable
testingurlfunction <- returnpolygonurl(latlongsac29,"drugs","2016-01")
#read the URL into a new table
crimeshere <- fromJSON(testingurlfunction)
```

## Gathering data on multiple neighbourhoods

Now that we've got it to work on one neighbourhood, we can expand to repeat this for *all* neighbourhoods in a particular force. Note that the `assign` function is used to create a new variable for each id code.

```{r}
boundaryurls <- c()
#trying to generate a URL like https://data.police.uk/api/greater-manchester/AC04/boundary
for (id in gmneighbourhoods$id){
  print(id)
  idurl <- paste("https://data.police.uk/api/greater-manchester/",id,"/boundary", sep="")
  print(idurl)
  #assign is another way of creating a variable, but you can use a variable *as* the name of the new variable
  #assign(paste(id,sep=""), fromJSON(idurl))
  #But rather than create a table variable to contain each neighbourhood, we can instead generate the URL we need directly, using the function we defined earlier
  holdthelatlongs <- fromJSON(idurl)
  holdtheurl <- returnpolygonurl(holdthelatlongs,"drugs","")
  boundaryurls <- c(boundaryurls,holdtheurl)
}
```

Again, this can be stored in a function so we can run it on another force. 

```{r}
#You need to first create an object containing all the neighbourhood ids, grabbed from the JSON
fetchboundaries <- function(force, crimetype){
  #trying to generate a URL like https://data.police.uk/api/greater-manchester/AC04/boundary
  for (id in force$id[1]){
    print(id)
    idurl <- paste("https://data.police.uk/api/",force,"/",id,"/boundary", sep="")
    print(idurl)
    #assign is another way of creating a variable, but you can use a variable *as* the name of the new variable
    holdtheurl <- returnpolygonurl(holdthelatlongs,"crimetype","")
    #assign is another way of creating a variable, but you can use a variable *as* the name of the new variable
    #assign(paste(id,sep=""), fromJSON(idurl))
    #But rather than create a table variable to contain each neighbourhood, we can instead generate the URL we need directly, using the function we defined earlier
    holdthelatlongs <- fromJSON(idurl)
    holdtheurl <- returnpolygonurl(holdthelatlongs,"drugs","")
    boundaryurls <- c(boundaryurls,holdtheurl)
    return(boundaryurls)
  }
}
```

Unfortunately, when we test this on a few areas, we find that most generate an error because some - in fact, the majority - have far too many lat/long pairs which results in a URL that the API cannot handle. Anything below 150 pairs seems OK, but over 170 is not.


## More

[More on getting polygons for neighbourhoods in this tutorial](http://jgardiner.co.uk/blog/police_data_api_in_leaflet_part2 )