---
title: "analysing spotify data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Grabbing the data

The data has already been scraped from the API using a Python script in Quickcode.io.The data can be queried using SQL with results provided in JSON. The URL for the result of `select * from swdata` (select all data) is [https://premium.scraperwiki.com/1c6u0ci/n4n1dvnblw9ieyh/sql/?q=select%20*%0Afrom%20swdata%0A](https://premium.scraperwiki.com/1c6u0ci/n4n1dvnblw9ieyh/sql/?q=select%20*%0Afrom%20swdata%0A).

To grab that we first need the `jsonlite` package. Then we can use the `fromJSON` command to grab it and put it in an object

```{r}
library("jsonlite")
spotifydata <- fromJSON("https://premium.scraperwiki.com/1c6u0ci/n4n1dvnblw9ieyh/sql/?q=select%20*%0Afrom%20swdata%0A")
```

## Exporting genres as CSV

To analyse the genres, I'm following the steps outlined in a [blog post by John Victor Anderson](http://johnvictoranderson.org/?p=115). 

First, we need to export the column of keywords:

```{r}
write(spotifydata$genres, 'genresastext.txt')
```

Now we re-import that data as a character object using `scan`:

```{r}
genres <- scan('genresastext.txt', what="char", sep=",")
# We convert all text to lower case to prevent any case sensitive issues with counting
genres <- tolower(genres)
```

We now need to put this through a series of conversions before we can generate a table:

```{r}
#create a list object by splitting 'genres' on comma
genres.split <- strsplit(genres, ",")
#create a vector object from that
genresvec <- unlist(genres.split)
#create a table from the vector
genrestable <- table(genresvec)
```

That table is enough to create a CSV from:

```{r}
write.csv(genrestable, 'genrecount.csv')
```



