---
title: "book notes"
author: "Paul Bradshaw"
date: "5 March 2017"
output: html_document
---

These are notes from the book [R for Data Science](http://r4ds.had.co.nz/).

Let's start by installing the tidyverse collection of packages, and activating it in our library.

```{r}
install.packages("tidyverse")
library(tidyverse)
```

And a few other packages that contain data we want to use:

```{r}
install.packages(c("nycflights13", "gapminder", "Lahman"))
```


## Story based inquiry, or hypothesis generation?

Hadley Wickham conveniently mirrors the division within journalism between hypothesis-driven investigation, and using data to generate leads:

> "It’s possible to divide data analysis into two camps: hypothesis generation and hypothesis confirmation (sometimes called confirmatory analysis)."

## Importing data

The tidyverse uses `read_csv` and `write_csv` as alternatives to R's own `read.csv` etc. However: "Note that the type information is lost when you save to csv ... you need to recreate the column specification every time you load in."

Alternatives include writing as RDS or feather (using the `feather` package).

[Other useful packages](http://r4ds.had.co.nz/data-import.html#other-types-of-data) include:

* `haven` (SPSS, Stata, and SAS)
* `readxl` (.xls and .xlsx).
* `DBI` "along with a database specific backend (e.g. RMySQL, RSQLite, RPostgreSQL etc) allows you to run SQL queries against a database and return a data frame."
* `jsonlite` for json
* `xml2` for XML. See: https://jennybc.github.io/purrr-tutorial/.
* `rio`

## Cleaning data: `parse_` functions

See the notebook in this repo.

## Combining data

See the notebook in this repo.