---
title: "summary tools"
author: "Paul Bradshaw"
date: "23 July 2018"
output: html_document
---

# Using the `summarytools` package

These are notes from following [the guide to the `summarytools` package](https://cran.r-project.org/web/packages/summarytools/vignettes/Introduction.html) on CRAN.

Let's install and activate the package first:

```{r}
install.packages("summarytools")
library(summarytools)
```

Now let's import some data to play with. I've got some food hygiene data:

```{r}
fooddata <- read.csv("bhamfooddata.csv")
#Clean up the column names which all begin with EstablishmentDetail...
colnames(fooddata) <-gsub("EstablishmentDetail...","",colnames(fooddata))
```

## The killer app: `dfSummary()`

The `dfSummary()` function provides a massive summary including histograms. Just run it and see:

```{r}
#You can run this without using view() but the results are prettier with it - you can see it in the Viewer panel on the right which should open when run.
view(dfSummary(iris))
```

If your data has anything non-numeric such as NA values then it will throw an error so you need to subset it to remove those first:

```{r}
foodnums <- data.frame(fooddata$Scores...Hygiene,fooddata$Scores...Structural,fooddata$Scores...ConfidenceInManagement)
foodnums.nonas <- subset(foodnums,foodnums$fooddata.Scores...Hygiene != "NA")
view(dfSummary(foodnums.nonas))
```


## Showing frequencies (pivot table - counts and % of column) using `freq()`

Now time to test out the functions in `summarytools`. First, `freq()`, which tells us what number and percentage of a column's cells contains each value:

```{r}
#We add the parameter style = "rmarkdown" which works nicely in a notebook. Other options are 'simple' or 'grid'
freq(fooddata$RatingValue, style = "rmarkdown")
```

Other parameters include:

* `report.nas = FALSE` to ignore `NA` values - we have 0 so it's not worth using, but it does remove those apparently duplicate extra 3 columns
* `totals = FALSE` to remove the total row

```{r}
#We add the parameter style = "rmarkdown" which works nicely in a notebook. Other options are 'simple' or 'grid'
freq(fooddata$RatingValue, style = "rmarkdown", report.nas = FALSE)
```

## Adding columns to our pivot table (crosstabulation) using `ctable()`

What if we want to add a column to our table to divide up the totals by another category? `ctable()` allows us to do that. 

We need to specify the row first and then the column like so:

`ctable(RatingValue,BusinessType)`

And then wrap that inside a `print() statement which in turn is contained in a `with()` function that specifies the table that those columns come from. The `print` function also has a `method =` parameter

```{r}
with(fooddata, print(ctable(RatingValue,BusinessType), method='render'))
```

We can exclude percentages by setting the `prop` parameter of `ctable` to "n":

```{r}
with(fooddata, 
     print(
  ctable(RatingValue,BusinessType, prop='n'), 
    method='render')
  )
```

Let's switch rows and columns - it's much easier to read:

```{r}
with(fooddata, 
     print(
  ctable(BusinessType,RatingValue, prop='n'), 
    method='render')
  )
```

## Descriptive statistics with `descr()`

The `descr()` function will generate means, medians, quartiles etc. for any numerical columns:

```{r}
descr(fooddata, style="rmarkdown")
```

Some useful parameters include the ability to specify *which* stats you want, and to transpose:

* `stats = c("mean", "sd", "min", "med", "max")`
* `transpose = TRUE`

```{r}
descr(fooddata, style="rmarkdown",transpose = TRUE,stats = c( "min", "Q1", "med", "Q3","max"))
```

What we see from this is that the individual hygiene score is typically better (3rd quartile is 5) than the other components (structural and management on 10 each - low scores are good in these component categories).

The results can be stored in a data object:

```{r}
stats <- descr(fooddata, style="rmarkdown",transpose = TRUE,stats = c( "min", "Q1", "med", "Q3","max"))
```

