---
title: "Using grep on MPs' expenses"
output: html_notebook
---

# Using grep and regex in R

First, let's get our data into R

```{r}
exesdata <- read.csv("Individual_claims_for_16_17.csv")
```

Now let's create a list of TRUE/FALSE values saying whether each row contains a particular pattern:

```{r}
#This looks for any match for .*Birmingham.* OR .*Sutton Coldfield.*
wmconstituency <- grepl(".*Birmingham,.*|.*Sutton Coldfield.*", exesdata$MP.s.Constituency)
```

Next, add it to the original data:

```{r}
exesdata$westmids <- wmconstituency
```

Now use that to create a subset:

```{r}
wmexes <- subset(exesdata, exesdata$westmids == TRUE)
```

## Advanced: doing this in fewer lines of code

In fact, instead of using 2 lines - one to create a TRUE/FALSE list, and another to add that to our data frame - we could have done both in one line like this:

```{r}
exesdata$westmids <- grepl(".*Birmingham,.*|.*Sutton Coldfield.*", exesdata$MP.s.Constituency)
```

And going further, we could actually *nest* part of the earlier line of code to do all 3 lines of the above code in one line: 

```{r}
westmidsonly <- subset(exesdata, grepl(".*Birmingham,.*|.*Sutton Coldfield.*", exesdata$MP.s.Constituency))
```

## Advanced: using `grep` to fetch values

The `grep` function will tell you *which* items in a vector (column) contain a match *or* - if `value=TRUE` is added - will return the text that matches. 

In the first case, that result is a list of numbers (indices indicating which row the matches are in). In the second case it is a list of strings, as shown below (I've added `head()` so you don't get lots of results):

```{r}
print("First the indexes")
head(grep("Const.*",westmidsonly$Expense.Type, value=FALSE))
print("Now with value set to false")
head(grep("Const.*",westmidsonly$Expense.Type, value=TRUE))
```

We can use `table()` to summarise the results instead and get a list of matching 'expense type' descriptions:

```{r}
table(grep("Const.*",westmidsonly$Expense.Type, value=TRUE))
```

This can be exported as a CSV to use in Excel as a lookup table:

```{r}
write.csv(table(grep("Const.*",westmidsonly$Expense.Type, value=TRUE)), "constituencytypes.csv")
```


